### 2018-07-01 paper/4-build.tex
More work on S4

More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using whatever ordering strategy
as per \S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The result will be a \hs{Bool} where \hs{True} indicates that the trace could be verified, or \hs{False} to indicate the key should be rebuilt.

@@

\subsubsection{Deterministic constructive traces}\label{sec-deterministic-constructive-traces}

If we assume all actions are \emph{deterministic} (specifically that given the same inputs they produce the same output) then we can make a further optimisation to arrive at deterministic constructive traces:

\begin{minted}{haskell}
data Trace k v = Trace
    { key     :: k
    , depends :: [(k, Hash (Tree (Hash v)))]
    , result  :: v
    }
\end{minted}

Instead of storing the hash of the immediate dependencies, a determinstic constructive trace only ever takes the hash of initial inputs. To form the hash of a dependency, it forms the tree of dependencies all the way to the inputs, and hashes that. As an example, if \hs{a} depends on the input \hs{b}, then the dependency value used for \hs{a} uses \hs{hash}~\hs{b}, \emph{but not} \hs{hash}~\hs{a}. As a consequence, given the hashes of all inputs, and dependency information (e.g. from an \hs{Applicative} constraint) it is possible to compute the dependency hash of \emph{any} node, without having built any intermediate nodes. This property is particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.

@@

In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
of traces~--~however, using a list necessarily means that finding the right trace
takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
a smarter representation, which we sketch below. Note that these implementations
do not avoid calls to \hs{compute}, merely overheads in the build system itself.

\begin{minted}{haskell}
data Trace k v = Trace
    { key     :: k
    , depends :: [(k, Hash v)]
    , result  :: Hash v }
\end{minted}

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} build system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result~--~i.e. \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result.
\item Many cloud build systems store hashes of values in the trace information, then have a separate ``blob server'' which associates hashes with their actual contents.
\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
\begin{minted}[xleftmargin=10pt]{haskell}
data Choice k v = Choice k (Map (Hash v) (Choice k v))
                | Result (Hash v)
\end{minted}
Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result. Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
\end{enumerate}

@@

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deterministic constructive traces\hspace{2mm}\hfill\S\ref{sec-deterministic-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{2mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-2mm}
\end{table}

>>>>

More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using an ordering strategy
as per \S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The result will be a \hs{Bool} where \hs{True} indicates that the trace could be verified, or \hs{False} to indicate the key should be rebuilt.

@@

\subsubsection{Deep constructive traces}\label{sec-deep-constructive-traces}

Constructive traces always verify keys by looking at their immediate dependencies, which must have first been brought up to date, meaning that the time to verify a key is based on the number of transitive dependencies. A \emph{deep} constructive trace optimises this process by only looking at the terminal input keys, ignoring any intermediate dependencies. The operations capturing this approach are the same as for constructive traces in~\S\ref{sec-constructive-traces}, but we use the names \hs{constructDCT} and \hs{recordDCT}, where the underlying \hs{DCT} state need only record information about hashes of inputs, not intermediate dependencies.

The deep constructive trace approach only works if the operations are \emph{deterministic}, otherwise inconsistencies arise (see~\S\ref{sec-non-determinism}). Concretely, if the result of building a key given a set of immediate dependencies can vary, then it is possible that key \hs{k} will have been computed locally, but something \emph{using} \hs{k} will have been taken as a constructive input using a different value for \hs{k}. If another rule combines those two values then two different values of \hs{k} will have been used in one rule, leading to problems such as link-errors for typical compilers.

While current build systems using deep constructive traces store only hashes of inputs, that is not a strict requirement, as the technique works equally well by skipping $n$ levels of dependency - where only inputs is the special case of $\infty$. Similarly, constructive traces can be described as taking the direct dependencies, and thus being deep constructive traces of $1$. A downside of deep constructive traces is that they can't support cutoff (\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On the otherhand, these traces are particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.

@@

In the examples above, we have used abstract types for the traces. Concretely, in our example code, they are all recorded as lists of:

\begin{minted}{haskell}
data Trace k h r = Trace
    { key     :: k
    , depends :: [(k, h)]
    , result  :: r }
\end{minted}

Where \hs{k} is the key type, \hs{h} is always \hs{Hash}~\hs{v} and \hs{r} is either \hs{Hash}~\hs{v} (verify traces) or \hs{v} (constructive traces). A real system is highly likey to use a more optimised implementation. Some of the most obvious optimisations are:

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} build system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result~--~i.e. \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result.
\item Many cloud build systems store hashes of values in the trace information, then have a separate ``blob server'' which associates hashes with their actual contents.
\end{enumerate}

@@

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{2mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-2mm}
\end{table}

### 2018-07-01 paper/5-implementations.tex
More work on S4

Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
constructive traces (\S\ref{sec-deterministic-constructive-traces}). As one can
see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
the current \hs{value}, since the \hs{constructDCT} query only needs to examine
the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.

>>>>

Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deep
constructive traces (\S\ref{sec-deep-constructive-traces}). As one can
see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
the current \hs{value}, since the \hs{constructDCT} query only needs to examine
the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.
